Initial environmental impacts of Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-08-13 manu-draft.Rmd
map fig added

@@

Ongoing threats and challenges to protecting water quality of Gulf Coast estuaries persist despite historical gains in environmental recovery.  Although point-source inputs of nutrient loads from wastewater treatment plants into Tampa Bay have been reduced, non-point sources from wastewater and stormwater runoff contribute to nutrient inputs, particularly during the rainy season from June to September [@Janicki08;@Janicki17].  Atmospheric deposition of nutrients from coal-based power production and automobile traffic further contribute about one-quarter of the total nitrogen inputs to the bay [@Poor13].  Climate change stressors, such as sea level rise, changing rainfall patterns, and temperature alterations, may further perturb ecosystem dynamics and assimilative capacity by reducing system resilience to nutrient inputs [@Sherwood14;@Burke17].  Many of these challenges are addressed by ongoing efforts of the US EPA National Estuary Program to implement a science-based resource management plan for the Bay[@Ohara17].  The Tampa Bay Estuary Program has been instrumental in coordinating efforts among local and regional partners to address legacy pollutants and current threats to the long-term protection of bay resources [@Greening14;@Greening16].

@@

The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer used in food production [@Henderson04].  By 2001, an estimated 40 million tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].  Currently, fourteen phosphogypsum stacks exist in the Tampa Bay watershed with no long-term plan for closure or disposal of waste to preven impacts to the environment.  The Piney Point facility located in Palmetto, Florida is a large phosphogypsum stack established in 2001 located less than two miles from the shore of Tampa Bay and near two aquatic preserves [@Henderson04].  Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, with oversight by the Florida Departmental of Environmental Protection (FLDEP).  Decreasing holding capacity of the ponds with seasonal rain events and storage of dredging material from nearby Port Manatee have contributed to degradation of the facility.  Discharges of wastewater from the stacks occurred in 2010 to nearby Bishops Harbor connected to Tampa Bay [@Garrett11].  Recently, FLDEP authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 29th, 2021 to release wastewater from the stacks directly into lower Tampa Bay to prevent catastrophic failure of the berms supporting the holding ponds.

@@

* 2021 timeline
* Historical context/baseline condition
* effluent characteristics
* water-column water quality changes
* red tide impacts
* Seagrass/macroalgae

@@

>>>>

library(here)

@@

Ongoing threats and challenges to protecting water quality of Gulf Coast estuaries persist despite historical gains in environmental recovery.  Although point-source inputs of nutrient loads from wastewater treatment plants into Tampa Bay have been reduced, non-point sources from wastewater and stormwater runoff contribute nutrients to the bay, particularly during the rainy season from June to September [@Janicki08;@Janicki17].  Atmospheric deposition of nutrients from coal-based power production and automobile traffic further contribute about one-quarter of the total nitrogen inputs to the bay [@Poor13].  Climate change stressors, such as sea level rise, changing rainfall patterns, and temperature alterations, may further perturb ecosystem dynamics and assimilative capacity by reducing system resilience to nutrient inputs [@Sherwood14;@Burke17].  Many of these challenges are addressed by ongoing efforts of the US EPA National Estuary Program to implement a science-based resource management plan for the Bay[@Ohara17].  The Tampa Bay Estuary Program has been instrumental in coordinating efforts among local and regional partners to address legacy pollutants and current threats to the long-term protection of bay resources [@Greening14;@Greening16].

@@

The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer used in food production [@Henderson04].  By 2001, an estimated 40 million tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].  Currently, fourteen phosphogypsum stacks exist in the Tampa Bay watershed with no long-term plan for closure or disposal of waste to preven impacts to the environment.  The Piney Point facility located in Palmetto, Florida is a large phosphogypsum stack established in 2001 located less than two miles from the shore of Tampa Bay and near two aquatic preserves [@Henderson04].  Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, with oversight by the Florida Departmental of Environmental Protection (FLDEP).  Decreasing holding capacity of the ponds with seasonal rain events and storage of dredging material from nearby Port Manatee have contributed to degradation of the facility.  Discharges of wastewater from the stacks occurred in 2010 to nearby Bishops Harbor connected to Tampa Bay [@Garrett11].  Recently, FLDEP authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 30th, 2021 to release wastewater from the stacks directly into lower Tampa Bay to prevent catastrophic failure of the berms supporting the holding ponds.

@@

* Figure \@ref(fig:timeline) timeline of events
* Fig 2 map, response and baseline data, w/ areas inset?
* Fig 3 Boxplots by week with +/-1sd for monthly long-term, or value median of long-term and monthly for seasonal eval
* Fig 4 seagrass and macroalgae, correlation matrix
* Fig 5 Red tide historic and current, include fish kill reports

* Table 1 Data collected
* Table 2 Stack/effluent characteristics, load estimates

@@

```{r timeline, fig.cap = 'Areas of interest for evaluating status and trends in response-based monitoring data and sample locations from March to July 2021 by monitoring data type in response to wastewater discharge from Piney Point.  Data types include algae sampling, seagrass and macroalgae, water quality (field-based and laboratory samples), and mixed monitoring (algae, seagrass and macroalgae, water quality).  Inset shows location of Tampa Bay on the Gulf coast of Florida, USA.'}
knitr::include_graphics(here('figs/map.jpeg'))
```

```{r map, fig.cap = 'Graphical timeline of events from the discharge of wastewater effluent at Piney Point starting Mrach 30th, 2021 through the end of July with the gradual decline of red tide in Tampa Bay.'}
knitr::include_graphics(here('figs/timeline.jpeg'))
```
